---
title: "Review problems"
author: "Michael Chow"
output: html_document
---

Data Structures and Types
=========================

1. Read in `dat1.Rdata`, which contains two columns of numbers (`a` and `b`). Add column `a` to `b`.

```{r}
load('dat1.Rdata')    # load variable named dat
dat                   # print it out

# add column a and b

as.integer(dat$a) + dat$b
```

2. How did the `data.frame` decide the names of its columns? How could you have it name them `a` and `b` without adding a new line of code? 

```{r}

dat <- data.frame(1:10, 11:20)

names(dat) <- c('a','b')
```

3. Write the `data.frame` below to a csv file using `write.csv`. Then, read it back in using `read.csv`. 
  What do you expect the column name to be when you read the data back in? Test your answer.
  What does this say about how R reads in data.frames?
```{r}

# new data frame
d <- data.frame(1:3)
names(d) <- "a 1"
names(d)

# write to csv below
write.csv(d, file='new_data.csv')

# read that csv in below
newdat <- read.csv('new_data.csv')

# what do you think the name of the column will be? Check it.
names(newdat)

# what about the data frame below?
data.frame("a 1"=1:3, "b"=2)

```


4. Without using the dollar sign (e.g. `dat$`) or blocks (e.g. `dat[]`), do the following..
  * paste columns `a` and `b` together
  * paste those columns again, except now create a third column named `c` with the results
  * do the same as above, but now add a fourth column as well (with anything you want)
  * get all rows where `b` is greater than 13

```{r}

dat <- data.frame(a = 1:10, b = 11:20)

# paste a and b together
with(dat, paste(a, b))

# make third column c
transform(dat, c = paste(a,b))

# make third and fourth columns
transform(dat, c = paste(a,b), d = 1)

# get rows where b is greater than 13
subset(dat, b > 13)

```

5. We want the vector `arr` below to sum to 5.
  * index `arr` to remove the values that are screwing up `sum` so that is sums correctly
  * try passing the `na.rm` argument to the sum function

```{r}
arr <- rep(c(1, NA), 5)

arr
sum(arr)     # should sum to 5

# subset arr below so that sum is correct
sum(arr[!is.na(arr)])

# try using na.rm argument of sum function
sum(arr, na.rm = TRUE)
```

6. We want the sum below to be 8. Re-assign two values of `a` to make this happen. What do you think will happen when you use the `na.rm` argument? Test your guess.

```{r}
a <- 3:8 / rep(-1:1, each=2)

# uncomment and modify the line below
a[is.infinite(a)] <- 0
sum(a)

# test na.rm argument below
sum(a, na.rm = TRUE)
```

7. Array indexing and counting.
  * Assign the indices in the array that correspond to "princeton" to the variable `indx`. Then, use this variable to get the values of "princeton" from the array.
  * Count the number of entries for each school

```{r}
arr <- sample(c('princeton', 'yale'), 10, replace=TRUE)

# assign indices to variable indx
indx <- which(arr == "princeton")


# get values of "princeton"
arr[indx]

table(arr)
```

8. Count the number of entries for each combination of school and gender (i.e. you will have 4 counts)

```{r}
dat <- data.frame(school=rep(c('princeton', 'yale', 'yale'), each=3),
           gender=rep(c('m', 'm', 'f'), 3))

table(dat[c('school', 'gender')])
```


Functions
=========

1. Note `runif(1)` yields a random value between 0 and 1.
  * pass `runif(1)` to a function that will repeat the value it produces 10 times
  * pass `runif(1)` to a function that will repeat `runif` to produce 10 random values

```{r}
# replace f with the correct function (you may need to give additional arguments)
rep(runif(1), 10)

# replace f for the second pard
replicate(10, runif(1))
```

2. `hist` is a simple function to make boxplots. 
  * Write a function, `makeBoxplot`, that takes a vector as it's first argument, adds 10 to the vector, then passes it to boxplot. Moreover, have that function use the ellipsis argument to pass other arguments to `hist`.
  * call `makeBoxplot(rnorm(100), main="I am a boxplot")`

```{r}
makeBoxplot <- function(x, ...){
  hist(x, ...)
}

makeBoxplot(rnorm(100), main="I am a boxplot")
```

3. In many cases, researchers will use a summary measure called standard error, which we will define as the standard deviation (`sd`), divided by the square root of the number of observations in a sample. Write a function, `se`, that returns the standard error of a vector of sample data. Modify the function so that optional arguments can be passed to the `sd` function within `se`.

```{r}
# use sd, length, and sqrt functions
se <- function(x, ...) {
  # complete this function
  sd(x, ...) / sqrt(length(x))
}
```

4. `tapply` is a function that breaks up a vector into groups, and applies a function to each group. You use it like `tapply(value_vector, group_vector, function, addition_arguments)`. Use the `se` function you created above to get the standard error of the values for each group. Be sure to pass the `na.rm` using `tapply(value_vector, group_vector, function, na.rm=TRUE)` argument to se.

```{r}
dat <- data.frame(value = c(rnorm(30, 10), rnorm(30), NA), 
                  group = c(rep(c('A', 'B'), each=30), NA))

# fill in tapply function
tapply(dat$value, dat$group, se, na.rm=TRUE)


```


Control Flow
============

1. For each number in the array below
  * if it is even, then print the square
  * if it is odd, then print the square-root
  
```{r}
arr <- 1:10
# put loop here

for (i in arr) {
  if (i %% 2 == 0) print(i^2)
  else print(sqrt(i))
}
```

2. Solve the above problem without using a for loop

```{r}

arr <- 1:10

# put solution here

```

3. [Fibonnaci Sequence](https://en.wikipedia.org/wiki/Fibonacci_number). Create a Fibonnaci Sequence that goes to 15 entries. By either..
  * creating an array of 15 zeroes, looping over it, and assigning the proper values to it
  * using the repeat statement rather than a for loop

```{r}



```


Errors
=============


1. What does this error mean?

```{r, error=TRUE}
a <- 3
x <- 5

eval(rm(list=tolower(LETTERS[20+4])))

a + x

```

2. What went wrong here? Could you create a similar error with another function?

```{r}

# What likely went wrong here?
dat <- read.csv('dat2.csv')
newdat <- subset(dat, a < 4)

# Similar error

```

3. What does this error mean?

```{r, error=TRUE}
load('dat3.Rdata')

df$b + 4
```

