---
title: "Basics of R homework"
output: html_document
---

Main problems 
-------------

1. Create a vector named "zip" with 5 elements, each one a digit of Princeton's zip code (08544).

``` {r}

# Replace this comment with your code


```

2.  Check the class of "zip".

``` {r}


```

3. Order of operations. Like a graphing calculator R performs arithmetic in a certain order. A common way to remember this order is PEMDAS: parentheses, exponent, multiplication, division, addition, subtraction. Using this, make the statements below all equal 1.

```{r}

1 + 2 / 3

2 ^ 3 - 3

3*2 / 9-3
```

4. Grab the first, third, and fifth elements of the vector below
```{r}

x <- 1:5

# uncomment, and fill in line below
# x[]

```


Exploring R functions
---------------------

### The `seq` function

Based on the output of the code below, how would you generate 6 evenly spaced numbers, from 1 to 10.

```{r}
seq(1,5,2)
seq(from=1, to=5, by=2)  # same as above
seq(from=1, to=5, length.out=3)
seq(from=1, to=5, length.out=2)
# try it here

```

How would you do it using the other option (either the `by` or `length.out` argument)?

```{r}



```

Note that `seq` can use arguments with decimals.
```{r}
seq(0, 1, by=.1)
```


### The `paste` and `rep` functions

**One interesting function that uses the concept of recycling is `paste`.**

```{r}
# LETTERS is a variable that is built in to R
LETTERS

LETTERS[1:4]
paste(LETTERS[1:4], 1:4)

paste(LETTERS[1:4], 1:3)

paste('the letter', LETTERS[1:4])

# note that by default it sepeartes entries with a space
# but you can specify a sep argument
paste('this letter is', LETTERS[1:4], sep=": ")
```

**Another which is often useful in conjunction with `paste` is `rep`.**

```{r}
ltrs <- LETTERS[1:4]
rep(ltrs, times=2)
rep(ltrs, each=2)
rep(ltrs, length.out=10)
```

How could you make the string counting up from 1 to 5 by .5 (that is, `1, 1.5, 2, ..., 5`), but label each number with it's whole part. For example, the first three entries should be: `"one: 1", "one: 1.5", "two: 2"`?

```{r}

# code here

```
