---
title: "Indexing and Summarizing Data - Homework "
output: html_document
---

Main Problems
-------------

1. Create a vector named "odd" with all odd numbers from 1 - 13.

``` {r}

```


2. Select the third element and raise it to the 5th power.

``` {r}

```


3. Create another vector named "even" with all even numbers from 2 - 14.

``` {r}

```


4. Create a data.frame named "numbers" with two columns: one named "apple" that contains your vector of odd numbers, and one named "orange" that contains your vector of even numbers.

``` {r}

```

5. Select all rows where "orange" is greater than 7.

``` {r}

```


6. Calculate the mean of the full "apple" column.

``` {r}

```

7. Without using the `$`, create a third column called `total` that is apples plus oranges

```{r}

```

### Recycling

The following might exhibit some surprising behavior, due to something called recycling..

```{r, error=TRUE}
x <- 1:4
x[c(TRUE, FALSE, FALSE, FALSE)]
x[c(TRUE, TRUE, TRUE, TRUE)]
x[TRUE]
x[c(TRUE, FALSE)]
x[c(TRUE, FALSE, FALSE)]
```

The data.frame function might offer some hints..

```{r}
data.frame(x = x, indx = c(TRUE, FALSE))
```

How can you get every 3 letters in the vector below (that is, `A, D, ..., Y`) using the fewest number of TRUE or FALSE possible. 
Note that the `LETTERS` variable is built in to R, and is a vector with A to Z.

```{r}
LETTERS[1:10]
# modify line below
LETTERS[c(TRUE)]

```