---
title: "Formulas"
output:
  revealjs::revealjs_presentation:
    theme: simple
    transition: slide
    center: true
    highlight: null
    includes:
      before_body: ../.style/prefix.html
---

```{r, echo=FALSE}

library(knitr)
opts_chunk$set(comment="",  warning=FALSE)

```

R Formulas
==========

What is a formula?
------------------

Formulas in R are created using the `~` character. The formula `A ~ B` is best read as "A by B".

The formula `A ~ B + C` can be read as "A by B **and C**".
This is well illustrated by the `boxplot` function.

Example of `A ~ B` using Boxplot
---------------------

ToothGrowth: **len**gth of teeth in guinea pigs, based on different **supp**lements, and **dose**s.

```{r}
boxplot(len ~ supp, data=ToothGrowth)
```

`A ~ B + C` w/ Boxplot
------------------------------------

```{r, fig.height=4.5}
boxplot(len ~ supp + dose, data=ToothGrowth)
```


What is the `data` argument?
=============================

Refresher, getting a column of data
------------------------------------

```{r}
small_tooth = ToothGrowth[1:5,]   # first 5 rows
small_tooth$len < 6
with(small_tooth, len < 6)
```

 The `data` argument
-----------------------------

```{r, fig.height=5}
plot(len ~ dose, data=ToothGrowth,
     col=supp)
```

Try it on the `iris` data
=========================

Questions (use formula)
------------------------

* Plot `Sepal.Length` by `Sepal.Width`
* Add color using the `Species` column to the plot above
* Make a boxplot of `Petal.Length` based on Species


Basic tests with R formulas
===========================

Overview
---------

* t-test
* one-way ANOVA
* two-way ANOVA

Data for examples
-----------------

We'll be looking more closely at the ToothGrowth data.

Use `?ToothGrowth` for details.

```{r}
table(ToothGrowth[,c('supp', 'dose')])
```

Boxplot of data
---------------

```{r}
boxplot(len ~ supp + dose, data=ToothGrowth)
```

t-test
-----------

Suppose we just want to compare these two groups...

```{r, fig.height=5}
boxplot(len ~ supp, data=ToothGrowth)
```

t-test (cont.)
--------------

```{r}
t.test(len ~ supp, data=ToothGrowth)
```

one-way ANOVA
-------------

```{r}
fit <- aov(len ~ dose, data=ToothGrowth)
summary(fit)
```

two-way ANOVA
-------------

```{r}
fit <- aov(len ~ supp*dose, data=ToothGrowth)
summary(fit)
```

two-way ANOVA, why the `*`?
----------------------------

```{r}
fit <- aov(len ~ supp + dose, data=ToothGrowth)
summary(fit)
```

**without the `*` we left out the interaction!**

